Gaussian Processes Autoregressive Models for the Disturbance Storm Time Index

### 2016-04-26 swsc.tex
Updates from ShareLaTeX

   \subtitle{Extracting predictive performance over and above persistence behavior}

@@

%Talk about Burton and friends
We focus on prediction of the hourly $D_{st}$ index which is an important problem because $D_{st}$ is a straight forward indicator of geomagnetic storms. More specifically we focus on the \emph{one step ahead} (OSA) prediction of $D_{st}$ because it is the first step towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.

The $D_{st}$ OSA prediction problem has been the subject of several modeling efforts so far, some of the earliest are by \citet{JGR:JGR10260} who outlined a prediction framework that has been used as a base line for many $D_{st}$ models. \citet{JGR:JGR10260} model $D_{st}(t)$ a function of time using an \emph{Ordinary Differential Equation} (ODE) which modeled it's rate of change as a combination of two terms: decay and injection $\frac{d D_{st}(t)}{dt} = Q(t) - \frac{D_{st}(t)}{\tau}$.

The \citet{JGR:JGR10260} model has proven to be very influential in the Space Weather research effort and many subsequent works like \citet{Wang:Dst}, \citet{JGRA:JGRA14856} modified the proposed ODE by defining expressions for the injection term $Q(t)$. More recently \citet{Ballatore2014} have tried to generate estimates for the injection and decay terms in Burton's equation.

%Talk about NARMAX Dst
Among the other major empirical models used to predict $D_{st}$ has been the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{JGRA:JGRA20661}, \citet{JGRA:JGRA50192} and \citet{balikhin:narmax}. The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations or monomials to include in the refined model using the \emph{error reduction ratio} (ERR) criterion. The parameters of the NARMAX OLS-ERR model are calculated using the standard \emph{ordinary least squares} (OLS) method hence these methods are also denoted by the acronym NARMAX OLS-ERR. The reader may refer to \citet{billings2013nonlinear} for a detailed exposition on the NARMAX methodology.

@@

%Assertion
It is with the above vision in mind, we propose \emph{Gaussian Process} models for OSA prediction of $D_{st}$. We use the results of \citet{Ji2012} as a starting point and compare our proposed models with the methods outlined in the same paper while using the performance of the \emph{Persistence} model as a litmus test.

% Talk about Gaussian Process
We use the hourly resolution Omni data set (see \cite{OmniPaper}) as extracted from NASA/GSFC's OMNI data set through OMNIWeb. The rest of the paper is organized as follows. Section \ref{sec:method} gives an introduction to the \emph{Gaussian Process} methodology for regression and the various issues surrounding their application. In section \ref{sec:osa} two different \emph{Gaussian Process} auto-regressive models are proposed and subsequently in sections \ref{sec:exp} and \ref{sec:res} bench marked against the \emph{Persistence} $D_{st}$ model as well as other models outlined in \citet{Ji2012}.

@@

\emph{Gaussian Processes} first appeared in machine learning research as the limiting case of Bayesian inference performed on neural networks with infinitely many neurons in the hidden layers \citet{Neal:1996:BLN:525544}. Although their inception in the machine learning community is recent, their origins can be traced back to the geo-statistics research community where they are known as \emph{Kriging} methods (\citet{krige1951statistical}). In pure mathematics area \emph{Gaussian Processes} have been studied extensively and their existence was first proven by Kolmogorov's extension theorem (\citet{tao2011introduction}). The reader is requested to refer to \cite{Rasmussen:2005:GPM:1162254} for an in depth treatment of Gaussian Processes in Machine Learning.

>>>>

   \subtitle{Attaining performance over and above persistence behavior}

@@

%Talk about Burton and friends
We focus on prediction of the hourly $D_{st}$ index which is an important problem because $D_{st}$ is a straightforward indicator of geomagnetic storms. More specifically we focus on the \emph{one step ahead} (OSA) prediction of $D_{st}$ because it is the first step towards building long term predictions of geomagnetic response of the Earth to changing space weather conditions.

The $D_{st}$ OSA prediction problem has been the subject of several modeling efforts so far, some of the earliest are by \citet{JGR:JGR10260} who outlined a prediction framework that has been used as a starting point for many $D_{st}$ models. \citet{JGR:JGR10260} model $D_{st}(t)$ a function of time using an \emph{Ordinary Differential Equation} (ODE) which expressed it's rate of change as a combination of two terms: decay and injection $\frac{d D_{st}(t)}{dt} = Q(t) - \frac{D_{st}(t)}{\tau}$.

The \citet{JGR:JGR10260} model has proven to be very influential in the Space Weather research effort and many subsequent works like \citet{Wang:Dst}, \citet{JGRA:JGRA14856} modified the proposed ODE by proposing expressions for the injection term $Q(t)$. More recently \citet{Ballatore2014} have tried to generate empirical estimates for the injection and decay terms in Burton's equation.

%Talk about NARMAX Dst
Among the other major empirical models used to predict $D_{st}$ has been the \emph{Nonlinear Auto-Regessive Moving Average with eXogenous inputs} (NARMAX) methodology developed in \citet{doi:10.1080/00207178908559767}, \citet{GRL:GRL13494}, \citet{GRL:GRL20944}, \citet{JGRA:JGRA18657}, \citet{balikhin:narmax}, \citet{JGRA:JGRA20661} and \citet{JGRA:JGRA50192}. The NARMAX methodology builds models by constructing polynomial expansions of inputs and determines the best combinations or monomials to include in the refined model using the \emph{error reduction ratio} (ERR) criterion. The parameters of the NARMAX OLS-ERR model are calculated using the standard \emph{ordinary least squares} (OLS) method hence these methods are also denoted by the acronym NARMAX OLS-ERR. The reader may refer to \citet{billings2013nonlinear} for a detailed exposition of the NARMAX methodology.

@@

%Assertion
It is with the above observation in mind, we propose \emph{Gaussian Process} models for OSA prediction of $D_{st}$. We use the results of \citet{Ji2012} as a starting point and compare our proposed models with the methods outlined in the same paper while using the performance of the \emph{Persistence} model as a litmus test.

% Talk about Gaussian Process
We use the hourly resolution Omni data set (see \cite{OmniPaper}) as extracted from NASA/GSFC's OMNI data set through OMNIWeb. The rest of the paper is organized as follows. Section \ref{sec:method} gives an introduction to the \emph{Gaussian Process} methodology for regression and the various issues surrounding their application. In section \ref{sec:osa} two different \emph{Gaussian Process} auto-regressive models are proposed and subsequently in sections \ref{sec:exp} and \ref{sec:res}, bench marked against the \emph{Persistence} $D_{st}$ model as well as other models outlined in \citet{Ji2012}.

@@

\emph{Gaussian Processes} first appeared in machine learning research in \citet{Neal:1996:BLN:525544}, as the limiting case of Bayesian inference performed on neural networks with infinitely many neurons in the hidden layers. Although their inception in the machine learning community is recent, their origins can be traced back to the geo-statistics research community where they are known as \emph{Kriging} methods (\citet{krige1951statistical}). In pure mathematics area \emph{Gaussian Processes} have been studied extensively and their existence was first proven by Kolmogorov's extension theorem (\citet{tao2011introduction}). The reader is requested to refer to \cite{Rasmussen:2005:GPM:1162254} for an in depth treatment of Gaussian Processes in machine learning.
